Predicting _C. difficile_ Infection Severity from the Taxonomic Composition of the Gut Microbiome

### 2023-06-05 paper/paper.qmd
Intro P1 polish & cite

_Clostridoides difficile_ infection (CDI) is the most common nosocomial
infection in the United States, and community-acquired cases are on the rise
[@magill_multistate_2014;@feuerstadt_burden_2023].
prevalence of severe cdi outcomes.
mortality ~9% [@napolitano_clostridium_2017].
antibiotics typical risk factor for cdi, but nonantibiotic medications can 
increase susceptibilty too [@tomkovich_osmotic_2021].
and non-nosocomial CDI on the rise.

Numerous studies indicate that the gut microbiome may play a role in _C. diff_ 
colonization, infection, and clearance.
Contribution of the gut microbiome.

prediction models based on EHR for whether infection occurs in the first place already in use [@otles_clostridioides_2023].
so how about predicting severity of infections to guide treatment.
also models on EHR to predict adverse outcomes (Li, Rao).
serum-based biomarker model (Dieterle mbio 2020).
OTUs vs EHRs to predict severity.
CDI severity prediction models could be deployed to screen patients at risk and
guide clinicians to consider prescribing a different course of treatment.
When paired with treatment options that may reduce risk of severity, deploying
prediction models can guide clinician decision-making to improve patient outcomes
while minimizing unnecessary harms.

A few ways to define CDI severity (@fig-flowchart).

The IDSA definition is known to be a poor predictor of adverse outcomes [@stevens_validation_2020],
however, it is easy to collect.

@@

We performed permutation feature importance to determine which OTUs contributed
the most to model performance.
An OTU was considered important if performance decreased when it was permuted in
at least 75% of the train/test splits, with greater differences in AUROC meaning
greater importance.
We plotted mean decrease in AUROC alongside log~10~-transformed mean
relative abundances for the top OTUs (@fig-features).
There is not always a clear pattern of increased or decreased relative abundance
for important OTUs in severe CDI cases, but all of the top 5 OTUs had an
increased mean relative abundance in severe cases relative to not severe cases.
_Enterococcus_ was the most important OTU, being significantly important for 
all models except for attributable severity on the full dataset.
_Staphylococcus_ was important for the pragmatic and all-cause definitions on
the full datasets, but not for models trained on the intersection dataset.
_Lactobacillus_ was important only for the all-cause definition on the intersection
dataset.
All remaining OTUs had differences in AUROC < 0.02 and were only significantly
important in one or two of the models at most.
TODO summarize which OTUs enriched in severe vs not severe.
TODO concluding sentence.

>>>>

_Clostridoides difficile_ infection (CDI) is the most common nosocomial
infection in the United States, and community-acquired cases are on the rise
[@magill_multistate_2014;@feuerstadt_burden_2023].
The classic CDI case typically occurs soon after antibiotic use, which perturbs the
protective gut microbiota and allows _C. diff_ to proliferate [@kelly_can_2012].
Non-antibiotic medications including proton-pump inhibitors and osmotic laxatives
have also been shown to increase CDI susceptibilty and inhibit clearance  
[@janarthanan_clostridium_2012;@tomkovich_osmotic_2021].
Diarrhea is the primary symptom, with some patients developing colitis, 
toxic megacolon, or requiring intensive care with an in-hospital mortality rate
of approximately 8-9% [@lucado_clostridium_2012;@kassam_clostridium_2016].
Furthermore, 5-20% of initial cases reoccur within 2-8 weeks, and recurrent
cases are associated with increased morbidity and mortality risk
[@napolitano_clostridium_2017;@kelly_can_2012].
CDI remains a significant burden on the US health care system with approximately
500,000 cases annually [@guh_trends_2020;@kwon_morbidity_2015].

Many factors influence CDI severity, 
however, there is not a consensus on how to
define severity [@ressler_defining_2021]. 
disease-related complications definition [@mcdonald_recommendations_2007].
tbh gut microbiome may only play a small role.
Numerous studies indicate that the gut microbiome may play a role in _C. diff_ 
colonization, infection, and clearance.
Contribution of the gut microbiome.

prediction models based on EHR for whether infection occurs in the first place already in use [@otles_clostridioides_2023].
so how about predicting severity of infections to guide treatment.
also models on EHR to predict adverse outcomes (Li, Rao ribotype).
serum-based biomarker model (Dieterle mbio 2020).
OTUs vs EHRs to predict severity.
CDI severity prediction models could be deployed to screen patients at risk and
guide clinicians to consider prescribing a different course of treatment.
When paired with treatment options that may reduce risk of severity, deploying
prediction models can guide clinician decision-making to improve patient outcomes
while minimizing unnecessary harms.

We chose to explore several different ways to define CDI severity (@fig-flowchart).
The IDSA definition is known to be a poor predictor of adverse outcomes [@stevens_validation_2020],
however, it is easy to collect.

@@

We performed permutation feature importance to determine which OTUs contributed
the most to model performance.
An OTU was considered important if performance decreased when it was permuted in
at least 75% of the train/test splits, with greater differences in AUROC meaning
greater importance.
We plotted mean decrease in AUROC alongside log~10~-transformed mean
relative abundances for the top OTUs (@fig-features).
_Enterococcus_ was the most important OTU, being significantly important for 
all models except for attributable severity on the full dataset.
_Staphylococcus_ was important for the pragmatic and all-cause definitions on
the full datasets, but not for models trained on the intersection dataset.
_Lactobacillus_ was important only for the all-cause definition on the intersection
dataset.
All remaining OTUs had differences in AUROC < 0.02 and were only significantly
important in one or two of the models at most.
There is not always a clear pattern of increased or decreased relative abundance
for important OTUs in severe CDI cases, but all of the top 5 OTUs had an
increased mean relative abundance in severe cases relative to not severe cases.
TODO concluding sentence.
